\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2019-05-09 paper/response_to_referee.txt

    The paper presents a method to estimate spectrophotometric parallaxes
    based only on input data, without adoption of physically motivated
    models and priors. The paper is written in an way that is strongly
    focused on the method itself, but leaves several complementary aspects
    that are certainly important to assess the reliability of the inferred
    parallaxes and, most importantly, distances. I recommend its
    publication only after such major issues are addressed.

    Major issues

    1 - The data-driven method described in the paper details out how
    spectrophotometric parallaxes are inferred. But the paper does not
    give the reader a clue about how spectrophotometric distances are
    estimated. The authors make it clear in Section 2 that they never
    invert *observed* parallaxes to estimate distances in their
    method. But are the spectrophotometric distances they infer (shown in
    Figures 5 and 6) just the inverse of the *inferred* parallaxes? If so,
    this should be made explicit in the text, along with some discussion
    about the problems (or lack of) that may arise from this
    inversion. Even though the spectrophotometric parallaxes themselves
    are always positive, a reader would like to be convinced that the
    distances shown in Figures 5 and 6 do not suffer from biasses caused
    by the asymmetric nature of the distances distribution that results
    from a symmetric error distribution in parallaxes (as stated by the
    likelihood assumption of the method). How do these biasses simply go
    away when the method's w(sp) +/- sigma(sp) are used? In particular, a
    negative parallax for many stars may be consistent with their w(sp)
    +/- sigma(sp).

These issues are subtle! We are assuming that the GAIA PARALLAXES have
noise that is symmetric in parallax space. In our data-driven model,
we use a linear combination of features to predict or estimate a LOG
PARALLAX, so if anything we are claiming that our estimates (not the
data, but our estimates based on the data) are symmetric in log
parallax, not parallax. But we aren't even really claiming that. The
issues are subtle. However, this does deserve some discussion and we
have added that to the captions of Figure XXX and YYY (formerly Figure
5 and 6), and in the last paragraph of the discussion, for good
emphasis [HOGG DO THIS AT "PLACEHOLDER A"].  The short answer is that
we just exponentiated the negative of our log-parallax estimates to
make those figures.

### 2019-05-13 paper/response_to_referee.txt

    The paper presents a method to estimate spectrophotometric parallaxes
    based only on input data, without adoption of physically motivated
    models and priors. The paper is written in an way that is strongly
    focused on the method itself, but leaves several complementary aspects
    that are certainly important to assess the reliability of the inferred
    parallaxes and, most importantly, distances. I recommend its
    publication only after such major issues are addressed.

    Major issues

    1 - The data-driven method described in the paper details out how
    spectrophotometric parallaxes are inferred. But the paper does not
    give the reader a clue about how spectrophotometric distances are
    estimated. The authors make it clear in Section 2 that they never
    invert *observed* parallaxes to estimate distances in their
    method. But are the spectrophotometric distances they infer (shown in
    Figures 5 and 6) just the inverse of the *inferred* parallaxes? If so,
    this should be made explicit in the text, along with some discussion
    about the problems (or lack of) that may arise from this
    inversion. Even though the spectrophotometric parallaxes themselves
    are always positive, a reader would like to be convinced that the
    distances shown in Figures 5 and 6 do not suffer from biasses caused
    by the asymmetric nature of the distances distribution that results
    from a symmetric error distribution in parallaxes (as stated by the
    likelihood assumption of the method). How do these biasses simply go
    away when the method's w(sp) +/- sigma(sp) are used? In particular, a
    negative parallax for many stars may be consistent with their w(sp)
    +/- sigma(sp).

These issues are subtle! We are assuming that the GAIA-MEASURED
PARALLAXES have noise that is symmetric in parallax space. In our
data-driven model, we use a linear combination of features to predict
or estimate a LOG PARALLAX, so if anything we are claiming that our
estimates (not the data, but our estimates based on the data) are
symmetric in log parallax, not parallax. But we aren't even really
claiming that. The issues are subtle. However, this does deserve some
discussion and we have added that to the captions of Figure XXX and
YYY (formerly Figure 5 and 6), and in the last paragraph of the
discussion, for good emphasis [HOGG DO THIS AT "PLACEHOLDER A"].  The
short answer is that we just exponentiated the negative of our
log-parallax estimates to make those figures.

@@

This is a great point. We obtained these two data sets (and the
Queiroz sample required some emails!) and we made comparisons. We
added a short paragraph about the results of these comparisons in
Section XXX. We did not feel that the comparisons merited supporting
Figures, but we do have such figures if the referee disagrees. HOGG DO THIS

During the long period in which we haven't responded to this report,
there was also a paper out by Leung & Bovy. They compare their results
to ours, and we also mention that in the newly added paragraph. HOGG DO THIS

    3 - The authors apply colour cuts to eliminate outliers in the APOGEE
    data, eqs 6 and 7. They also apply a cut in parallax error, eq. 8.,
    along with other quality criteria, to define the training set. So, the
    training set parallax uncertainty distribution is truncated. How do
    these cuts affect the method and its assumptions? In particular, it is
    important to reconcile these selection criteria for the training set
    with the statements presented in the Introduction (page 3):
    "... regressions can be strongly biased if we have a bad selection
    function or data censoring. This means that we cannot apply cuts to
    the Gaia astrometry or signal-to-noise unless those cuts are correctly
    accounted for in our generative model for the data sample. Below, we
    will make no such cuts."  It is in fact clear from Figure 1 that the
    training set is well under represented of low log g stars with respect
    to the parent sample. This already jeopardizes the representativeness
    assumption somewhat.  One useful assessment of the effects of data
    censoring would be to show the spectral S/N or parallax distributions
    of the training data in comparison to that of the entire sample that
    the method is applied to.

This point is important. The Gaia parallax uncertainties are based on
a spacecraft model that is independent of the measurements themselves.
Therefore, cuts on parallax uncertainties are not cuts on parallax and
thus are not subject to the strong censure that we give to such cuts!
But the referee is correct that we have been very unclear about this
and we have attempted to fix this in the relevant paragraph towards the
end of Section 2. This has improved the paper.

@@

Absolutely! Actually we had a section about this but dropped it just
before submission. One of the reasons we dropped it is because the
story is somewhat complicated. But you are right that this must be
there, and we have added it back in, at the end of Section YYY. The
short story (which you can read in the updated text) is that the
photometric part of the model is interpretable and stable, but the
spectroscopic part is less stable in its internals (while remaining
very stable in its predictions in the data space). We don't fully
understand this, but it connects to known issues in machine learning
more generally, which we point out in the text. HOGG DO THIS

    Other issues:

    Introduction, 1st par.: could the authors say a few words about the
    importance of making "precise kinematic and element-abundance maps of
    the Milky Way disk out to large heliocentric radii..."?

Good point! The paper now opens with the big question and flows to the
spectroscopy and precision. This change affects the first paragraph of
the paper and changes how it starts.

@@

    Section 4: "The hope of the model assumptions is that the Training Set
    will contain sufficient dust variation that the model will naturally
    learn the dust corrections and extrapolate acceptably.” Is this hope
    checked in any way?

Hahaha no. Okay fine. We now comment on this point in Section YYY
where we deliver the interpretation of the photometric part of the
model. HOGG DO THIS.

    Section 4: The parallax offset w_0 is a contentious issue and it is
    important to quantify its effect on the final spectrophotometric
    parallaxes.

The effect is very weak; we don't find that the change from 0.029 to 0.048
makes much difference to our results except at very small parallaxes. And
our 0.048 is not preferred over 0.029 at large confidence. The main reason
it has little effect is that it effectively moves the training data only a
tiny fraction of a sigma. We added two sentences to this effect in the relevant
place.

### 2019-05-13 paper/sp.tex
more referee items

%% This file is part of the spectroscopic_parallax project
%% Copyright 2018 the authors.

@@

If we want to make precise kinematic and element-abundance maps
of the Milky Way disk out to large heliocentric radii,
and especially through the extinction to the other side of the Galactic Center, we will
need to make use of luminous red giants, and we will need to observe in
the infrared.
These arguments motivate the \apogee\ (\citealt{apogee}),
\apogee-2, and \sdssv\ (\citealt{sdssv})
projects, which take spectra in the infrared, and deliver detailed abundances
along the entire red-giant branch up to the tip.
The maps made with these data will reveal the dynamics of the Milky Way and its disk,
and show us how and where stars form, and how they migrate around the
disk with time.

@@

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design.
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise; we include
training parallax measurements even if they are negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

In addition to making the above assumptions, we also avoid various practices
with the data that are tempting but lead to important biases.
For example, we never cut on \gaia\ parallax or parallax signal-to-noise.
The common practices of cutting to parallaxes that are good to 20~percent
(see, for example, \citealt{trick, helmi}),
or Catalog parallaxes that are positive, or parallaxes that are smaller or
larger than something, are all practices that will bias results on stellar
collections.
That is, if you cut on parallax or parallax signal-to-noise and you subsequently
take an average of parallaxes for some population, or perform a regression (as we
do here), the results of that average or regression will be strongly biased.
We never cut on parallax or parallax signal-to-noise.
By using a justifiable likelihood function (the ``likelihood'' assumption above),
we can use all of the \gaia\ parallaxes without the low signal-to-noise and
negative parallaxes causing trouble for our regressions, and without the biases
that enter when cuts are made.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended all-sky-average value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
It also happens to be in the range of other values found in the literature (\citealt{arenou, zinn}).
Of course we really expect the offset to be a function of sky position and color (at least),
so the offset we find is not recommended for further use as any kind of universal value.
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.

Comparison of the top and bottom panels of {\figurename}~\ref{fig:samples} shows
that there is more color range in the Parent Sample than in the Training Set.
This challenges the representativeness assumption in \sectionname~\ref{sec:assumptions}.
The principal reason for the difference is that the \gaia\ quality cuts
exclude stars preferentially from crowded regions, which also tend to be the
most dust-obscured.
The hope of the model assumptions is that the Training Set will contain sufficient
dust variation that the model will naturally learn the dust corrections and extrapolate
acceptably.

>>>>

%% This file is part of the spectroscopic_parallax project
%% Copyright 2018, 2019 the authors.

@@

The Milky Way disk is a beautiful object of study for cosmology and galaxy evolution.
Its evolution involves gas accretion and cooling, star formation, and orbital dynamics.
To understand the details of this story, we need precise measurements of the
stars and their properties over a large part of the disk, not just
very near the Sun.
Since we live in the middle of this disk, and the disk is large and
dusty, these measurements require infrared spectroscopy of luminous
giants.
This motivates the \apogee\ (\citealt{apogee}),
\apogee-2, and \sdssv\ (\citealt{sdssv})
projects, which take spectra in the infrared, and deliver detailed abundances
along the entire red-giant branch up to the tip.
The maps made with these data will reveal the dynamics of the Milky Way and its disk,
and show us how and where stars form, and how they migrate around the
disk with time.

@@

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design.
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise; we include
training parallax measurements even if they are negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts on astrometric measurements
and we keep an umodified, justified likelihood function
is unbiased, and simple.

@@

In addition to making the above assumptions, we also avoid various practices
with the data that are tempting but lead to important biases.
The common practices of cutting to parallaxes that are good to 20~percent
(see, for example, \citealt{trick, helmi}),
or Catalog parallaxes that are positive, or parallaxes that are smaller or
larger than something, are all practices that will bias results on stellar
collections.
That is, if you cut on parallax or parallax signal-to-noise and you subsequently
take an average of parallaxes for some population, or perform a regression (as we
do here), the results of that average or regression will be strongly biased.
We never cut on parallax or parallax signal-to-noise.
We will (below) cut on \gaia\ \emph{uncertainty estimates} on parallax.
These uncertainty estimates are a function only of spacecraft orientation history
(scanning law) and telemetry; they are not a function of the magnitudes of the
parallaxes or their measurements.
For these reasons, the uncertainty estimates on the parallaxes can be used to
make sample cuts (and we do use them).
These cuts will not bias our regressions.
By using a justifiable likelihood function (the ``likelihood'' assumption above),
we can use all of the \gaia\ parallaxes without fearing that the low signal-to-noise and
negative parallaxes causing trouble for our regressions, and without any of the biases
that enter when cuts are made.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended all-sky-average value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
It also happens to be in the range of other values found in the literature (\citealt{arenou, zinn}).
Of course we really expect the offset to be a function of sky position and color (at least),
so the offset we find is not recommended for further use as any kind of universal value.
The spectrophotometric parallaxes we generate
do not depend strongly on this choice.
It does not have a large effect because it moves the training data by only a tiny
fraction of the root-variance of the uncertainties.
It is also the case (for the same reason) that we do not strongly prefer (in a statistical sense)
the $\varpi_0 = 0.048$\,mas value over the \gaia\ value of 0.029\,mas.

The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.
Comparison of the top and bottom panels of {\figurename}~\ref{fig:samples} shows
that there is more color range in the Parent Sample than in the Training Set.
This challenges the representativeness assumption in \sectionname~\ref{sec:assumptions}.
The principal reason for the difference is that the \gaia\ quality cuts
exclude stars preferentially from crowded regions, which also tend to be the
most dust-obscured.
The hope of the model assumptions is that the Training Set will contain sufficient
dust variation that the model will naturally learn the dust corrections and extrapolate
acceptably.
